---
title: "Convergence club analysis in R"
author: "Carlos Mendez"
output:
  html_notebook:
    code_folding: show
    highlight: monochrome
    number_sections: yes
    theme: cosmo
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
      smooth_scroll: no
  html_document:
    code_download: true
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 4
    number_sections: true
    code_folding: "show"
    theme: "cosmo"
    highlight: "monochrome"
  pdf_document: default
  word_document: default
  github_document: default
bibliography: biblio.bib
---

<style>
h1.title {font-size: 18pt; color: DarkBlue;} 
body, h1, h2, h3, h4 {font-family: "Palatino", serif;}
body {font-size: 12pt;}
/* Headers */
h1,h2,h3,h4,h5,h6{font-size: 14pt; color: #00008B;}
body {color: #333333;}
a, a:hover {color: #8B3A62;}
pre {font-size: 12px;}
</style>

Suggested citation: 

> Mendez C. (2020).  Convergence club analysis in R. R Studio/RPubs. Available at <https://rpubs.com/quarcs-lab/tutorial-convergence-clubs>

This work is licensed under the Creative Commons Attribution-Non Commercial-Share Alike 4.0 International License. 


Acknowledgment:

Material adapted from multiple sources, in particular [Sichera and Pizzuto (2019)]https://journal.r-project.org/archive/2019/RJ-2019-021/RJ-2019-021.pdf)

# Libraries

## Install libraries

```{r eval=F, include=TRUE}
install.packages("devtools")
devtools::install_github("rhobis/ConvergenceClubs")
packageVersion("ConvergenceClubs") 
```

```{r eval=F, include=TRUE}
install.packages("tidyverse")
packageVersion("tidyverse") 
```


## Load librries

```{r message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ConvergenceClubs)
library(tidyverse)


# Change the presentation of decimal numbers to 4 and avoid scientific notation
options(prompt="R> ", digits=4, scipen=7)
```

# Tutorial objectives

- Import `.csv` from github



# Replication files

- If you are a member of the [QuaRCS lab](https://quarcs-lab.rbind.io/), you can run this tutorial in [R Studio Cloud](https://rstudio.cloud/spaces/15597/project/965714) and access the files in the following [Github Repository](https://github.com/quarcs-lab/ConvergenceClubs)

# Preliminary material

# Import data




# References and tutorials

- [Sichera, R. and Pizzuto, P., 2019. ConvergenceClubs: A Package for Performing the Phillips and Sul's Club Convergence Clustering Procedure. The R Journal](https://journal.r-project.org/archive/2019/RJ-2019-021/RJ-2019-021.pdf). 
- <https://rhobis.rbind.io/convergenceclubs/index.html>
- <https://rhobis.rbind.io/convergenceclubs/reference/index.html>
- <https://cran.r-project.org/web/packages/ConvergenceClubs>
- <https://github.com/rhobis/ConvergenceClubs>
- <https://rhobis.rbind.io/blog/convergenceclubs-package-release>
- <https://rpubs.com/econdata777/ConvergenceClub-package>




END
